# Regression Analysis of Household Trip Rates
=============================================

## Description
We develop multiple regression models to estimate household trip rates.

* We focus on four different types of home-based trips, including home-based work (HBW), home-based shopping (HBS), home-based recreation (HBR), and home-based other (HBO).
* The dependent variable is the number of linked trips, not place-based trips, at the houehold level. For example, in the Year 2011 Oregon Travel and Activity Survey, which is a place-based survey, a transit trip is coded with 3 trips, including transit access, in transit, and transit egress. In our study, we use linked trips; that is, we convert the transit trip from 3 trips to 1 trip.
* We have two sets of independent variables: sociodemographic characteristics of households (SD) and built environment characteristics of household's home census block groups (BE).
    + the SD variables: household size, number of workers, number of students, household income, etc.
    + the BE variables: EPA's Smart Location Database (Density (D1), Diversity (D2), Design (D3), Transit (D4), and Destination Accessibility (D5)

### Load data from the PostgreSQL database
* the activity file of the 2011 OTAS
* the household file of the 2011 OTAS
    + households' home census block groups (hcbg) were spatialled joined
* EPA's Smart Location Database
    + downloaded from [Smart Location Mapping](http://www.epa.gov/smartgrowth/smartlocationdatabase.htm) for our study area, followed by converting the Coordinate Reference System to EPSG 2913
    + for a full description of all variablels, please refer to [its Users' Guide](http://www.epa.gov/smartgrowth/pdf/sld_userguide.pdf)
* we combined the household file and the Smart Location Database, which is hh_sld
```{r, echo=FALSE, message=FALSE}
setwd("/Users/Potenza/Documents/02_Projects/TCI")
library(RPostgreSQL)
library(pscl)
library(memisc)
conn <- dbConnect(PostgreSQL(), host="", user="", password="", dbname="portland")
hh <- dbReadTable(conn, c("ohas_v2", "household")) # household file from the 2011 OTAS
sld <- dbReadTable(conn, c("ohas_v2", "sld_metro")) # EPA's Smart Location Database
hh_sld <- dbReadTable(conn, c("ohas_v2","hh_sld")) # a spatially joined table of household and Smart Location database 
act <- dbReadTable(conn, c("ohas_v2", "activity")) # activity file from the 2011 OTAS
act <- with(act, act[order(sampn,perno,plano),])
act$actNo <- with(act, ave(plano, sampn,perno, FUN=seq))
act$maxAct <- with(act, ave(actNo, sampn,perno, FUN=length))
```

### Create a dependent variable for each home-based trip purpose : number of linked trips per household
* hba (home-based all); all home-based trips
* hbw (home-based work); work, not including all other activities at work and work related activities
* hbs (home-based shopping); both routine and special shopping
* hbr (home-based recreation); both outdoor and indoor recreation
* hbo (home-based other); other home-based trips
```{r}
# create a new data set without the Change-of-Mode activities in the middle of the day to use only linked trips
linked <- subset(act, !(tpurp==7 & actNo!=1 & actNo!=maxAct))
# update actNo and maxAct
linked$actNo <- with(linked, ave(plano, sampn,perno, FUN=seq))
linked$maxAct <- with(linked, ave(actNo, sampn,perno, FUN=length))
# update lastaggact
linked$lastaggact <- c(NA, linked$thisaggact[-length(linked$thisaggact)])
linked$lastaggact[linked$actNo==1] <- NA
# update last_tpurp
linked$last_tpurp <- c(NA, linked$tpurp[-length(linked$tpurp)])
linked$last_tpurp[linked$actNo==1] <- NA
# identity trip purposes
linked$h2o <- with(linked, ifelse(actNo!=1
                                  & (lastaggact=='Home'|lastaggact=='WorkAtHome')
                                  & (thisaggact!='Home'&thisaggact!='WorkAtHome'), 1, 0))
linked$o2h <- with(linked, ifelse(actNo!=1
                                  & (lastaggact!='Home'&lastaggact!='WorkAtHome')
                                  & (thisaggact=='Home'|thisaggact=='WorkAtHome'), 1, 0))
linked$hba <- with(linked, ifelse(h2o==1|o2h==1, 1, 0))
linked$hbw <- with(linked, ifelse((h2o==1&tpurp==3)|(o2h==1&last_tpurp==3), 1, 0))
linked$hbs <- with(linked, ifelse((h2o==1&(tpurp==13|tpurp==14))|(o2h==1&(last_tpurp==13|last_tpurp==14)), 1, 0))
linked$hbr <- with(linked, ifelse((h2o==1&(tpurp==20|tpurp==21))|(o2h==1&(last_tpurp==20|last_tpurp==21)), 1, 0))
linked$hbo <- with(linked, ifelse((h2o==1&(tpurp!=3&tpurp!=13&tpurp!=14&tpurp!=20&tpurp!=21))
                                  |(o2h==1&(last_tpurp!=3&last_tpurp!=13&last_tpurp!=14&last_tpurp!=20&last_tpurp!=21)), 1, 0))
# calculate the number of linked trips by home-based trip purposes at the household level
hba <- setNames(aggregate(linked[,"hba"], by=list(linked$sampn), "sum"), c("sampn", "numtrips"))
hbw <- setNames(aggregate(linked[,"hbw"], by=list(linked$sampn), "sum"), c("sampn", "numtrips"))
hbs <- setNames(aggregate(linked[,"hbs"], by=list(linked$sampn), "sum"), c("sampn", "numtrips"))
hbr <- setNames(aggregate(linked[,"hbr"], by=list(linked$sampn), "sum"), c("sampn", "numtrips"))
hbo <- setNames(aggregate(linked[,"hbo"], by=list(linked$sampn), "sum"), c("sampn", "numtrips"))
```

### Add independent variables
* the sociodemographic characteristics of households
* the built environment characteristics of household home census block groups
```{r, echo=FALSE}
# merge dependent variables with independent variables
hba <- merge(hba, hh_sld, by="sampn")
hbw <- merge(hbw, hh_sld, by="sampn")
hbs <- merge(hbs, hh_sld, by="sampn")
hbr <- merge(hbr, hh_sld, by="sampn")
hbo <- merge(hbo, hh_sld, by="sampn")
pairs(~numtrips+hhsiz+hhwrk+hhveh, data=hbw)
pairs(~numtrips+d1a+d2b_e5mix+d3a, data=hbw)
```

### Develop regression models to estimate household trip rates
```{r, echo=FALSE}
# only with sociodemographic variables
f1 <- as.formula(numtrips ~ hhsiz + hhwrk + hhstu + hhveh + hhlic + flexc)
f1.hba <- lm(f1, data=hba)
f1.hbw <- lm(f1, data=hbw)
f1.hbw_workers <- lm(f1, data=hbw[which(hbw$hhwrk!=0),])
f1.hbs <- lm(f1, data=hbs)
f1.hbr <- lm(f1, data=hbr)
f1.hbo <- lm(f1, data=hbo)
f1.outputs <- mtable("HBA"=f1.hba,"HBW"=f1.hbw,"HBW_workers"=f1.hbw_workers,"HBS"=f1.hbs,"HBR"=f1.hbr,"HBO"=f1.hbo)
print(f1.outputs)
# add built environment variables
f2 <- update(f1, ~ . + d1a + d1b + d1c5_ret10
                     + d2a_jphh + d2b_e5mix + d2a_ephhm
                     + d3a + d3b)
f2.hba <- lm(f2, data=hba)
f2.hbw <- lm(f2, data=hbw)
f2.hbw_workers <- lm(f2, data=hbw[which(hbw$hhwrk!=0),])
f2.hbs <- lm(f2, data=hbs)
f2.hbr <- lm(f2, data=hbr)
f2.hbo <- lm(f2, data=hbo)
f2.outputs <- mtable("HBA"=f2.hba,"HBW"=f2.hbw,"HBW_workers"=f2.hbw_workers,"HBS"=f2.hbs,"HBR"=f2.hbr,"HBO"=f2.hbo)
print(f2.outputs)
```

### Develop zero-inflated negative binomial (zinb) models
```{r}
zinb.hbw <- zeroinfl(numtrips ~ hhsiz + hhstu + hhveh + hhlic + flexc +
                 d1a + d1b + d1c5_ret10 + d2a_jphh + d2b_e5mix + d2a_ephhm + d3a + d3b | hhwrk,
                 data=hbw, dist="negbin", EM=TRUE)
summary(zinb.hbw)
```
