---
title: "Comparison of OHAS and NHTS Trip Costs"
author: "HuajieYang and Liming Wang, Ph.D."
date: "9/9/2015"
output: html_document
---

### Overview 
This document compares the results of OHAS and NHTS.

```{r, echo=FALSE}
#TODO: add execution steps to create charts/tables from raw input data
setwd("~/tci")
source("code/misc/NHTS09.R")
source("code/OHAS/start.R")

  # make names for household income groups, trip purpose and calculation method
  IcNames <- c("Low Income", "Mid Income", "High Income")
  Ic <- c("lowInc", "midInc", "highInc")
  names(IcNames) <- Ic
  
  PrNames <- c("Work", "Shopping", "Recreation", "Other")
  Pr <- c("hbw", "hbs", "hbr", "hbo")
  names(PrNames) <- Pr
  
  CmNames <- c("mintcost", "avgtcost", "maxtcost")
  Cm <- c("min", "avg", "max")
  names(CmNames) <- Cm

```

### OHAS
```{r, echo=FALSE}
setwd("~/tci")
load("output/OHAS/tcost.RData")
```

**OHAS data description**
```{r}
# Number of household
nrow(tcost.hh)

# Number of household by household size
tcost.hh.hhsiz.freq <- tcost.hh %>%
                 group_by(HHSIZ) %>%
                 summarise(freq=n())
tcost.hh.hhsiz.freq

# Number of trips by income groups and and trip purposes 
tcost.hh.tpurp.inc.freq <- tcost.trip %>%
                           group_by(TripPurpose, inc.level) %>%
                           summarise(freq=n()) %>%
                           as.data.frame()
tcost.hh.tpurp.inc.freq
```


**OHAS: density_tcost.hh_by_inc.png**

```{r}
# Plot trip cost by income groups 
pden.inc <- ggplot(tcost.hh, aes(x = tcost, colour=inc.level, group=inc.level)) +
            geom_density(fill=NA, size=1) + labs(x="Travel Costs (minutes)") + xlim(0, 360) +
            scale_colour_discrete(name = 'Income Level') + 
  ggtitle("Household-level travel cost by household size") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))

pden.inc

```


**OHAS: density_tcost.hh_by_hhsiz.png**
```{r}
tcost.hh <- tcost.hh %>% 
  mutate(hhsiz.cat=cut(HHSIZ,
                       breaks=c(1, 2, 3, 4, 9),
                       labels=c("1", "2", "3", "4+"),   #allow alternative household grouping
                       include.lowest=T, right=F
  ))
pden.hhsiz <- ggplot(tcost.hh, aes(x = tcost, colour=hhsiz.cat, group=hhsiz.cat)) +
  geom_density(fill=NA, size=1) + labs(x="Travel Costs (minutes)") + xlim(0, 360) +
  scale_colour_discrete(name = 'Household Size') + 
  ggtitle("Household-level travel cost by household size") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))
pden.hhsiz
```

**OHAS: boxplot_tcost.hh_by_tpurp.inc.png**

```{r}
tcost.hh.tpurp <- tcost.hh.tpurp %>% 
  mutate(inc.level = factor(inc.level, levels=Ic, labels=c("Low Inc", "Mid Inc", "High Inc")),
         TripPurpose = factor(TripPurpose, levels=Pr, labels=c("HBW", "HB Shopping", "HB Recreation", "HB Other"))
  )

boxp.tpurp_inc <- ggplot(tcost.hh.tpurp, aes(x=TripPurpose, y=tcost, fill=inc.level)) +
  geom_boxplot() + labs(y="Generalized Travel Costs (minutes)") + xlab("Trip Purpose") + ylim(0, 250)  +
  scale_fill_discrete(name = 'Income Level') + 
  ggtitle("Household-level travel cost by trip purposes income groups") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))
boxp.tpurp_inc
```

**OHAS: lineplot_tcost.hh_by_tpurp.inc.png**
```{r}
t.lc <- ggplot(tcost.tpurp.inc, aes(x = inc.level, y = tcost.wtavg, colour=TripPurpose, group=TripPurpose))
t.lc + geom_line(fill=NA, size=1) + labs(x="Income Level") + labs(y="Travel Costs (minutes)") + ylim(0, 120) +
  ggtitle("Trip-level travel cost by trip purposes income groups") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))
```

### Portland
```{r, echo=FALSE}
setwd("~/tci")
load("output/NHTS09/Portland/tcost.RData")
```


**Portland data description**
```{r}
# Number of household
nrow(tcost.hh)

# Number of household by household size
tcost.hh.hhsiz.freq <- tcost.hh %>%
                 group_by(HHSIZ) %>%
                 summarise(freq=n())
tcost.hh.hhsiz.freq
# Number of trips by income groups and and trip purposes 
tcost.hh.tpurp.inc.freq <- tcost.trip %>%
                           group_by(TripPurpose, inc.level) %>%
                           summarise(freq=n())%>%
                           as.data.frame()
tcost.hh.tpurp.inc.freq
```



**Portland: density_tcost.hh_by_inc.png**

```{r}
# Plot trip cost by income groups 
pden.inc <- ggplot(tcost.hh, aes(x = tcost, colour=inc.level, group=inc.level)) +
            geom_density(fill=NA, size=1) + labs(x="Travel Costs (minutes)") + xlim(0, 360) +
            scale_colour_discrete(name = 'Income Level') + 
  ggtitle("Household-level travel cost by household size") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))

pden.inc

```


**Portland: density_tcost.hh_by_hhsiz.png**
```{r}
tcost.hh <- tcost.hh %>% 
  mutate(hhsiz.cat=cut(HHSIZ,
                       breaks=c(1, 2, 3, 4, 9),
                       labels=c("1", "2", "3", "4+"),   #allow alternative household grouping
                       include.lowest=T, right=F
  ))
pden.hhsiz <- ggplot(tcost.hh, aes(x = tcost, colour=hhsiz.cat, group=hhsiz.cat)) +
  geom_density(fill=NA, size=1) + labs(x="Travel Costs (minutes)") + xlim(0, 360) +
  scale_colour_discrete(name = 'Household Size') + 
  ggtitle("Household-level travel cost by household size") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))
pden.hhsiz
```

**Portland: boxplot_tcost.hh_by_tpurp.inc.png**

```{r}
tcost.hh.tpurp <- tcost.hh.tpurp %>% 
  mutate(inc.level = factor(inc.level, levels=Ic, labels=c("Low Inc", "Mid Inc", "High Inc")),
         TripPurpose = factor(TripPurpose, levels=Pr, labels=c("HBW", "HB Shopping", "HB Recreation", "HB Other"))
  )

boxp.tpurp_inc <- ggplot(tcost.hh.tpurp, aes(x=TripPurpose, y=tcost, fill=inc.level)) +
  geom_boxplot() + labs(y="Generalized Travel Costs (minutes)") + xlab("Trip Purpose") + ylim(0, 250)  +
  scale_fill_discrete(name = 'Income Level') + 
  ggtitle("Household-level travel cost by trip purposes income groups") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))
boxp.tpurp_inc
```

**Portland: lineplot_tcost.hh_by_tpurp.inc.png**
```{r}
t.lc <- ggplot(tcost.tpurp.inc, aes(x = inc.level, y = tcost.wtavg, colour=TripPurpose, group=TripPurpose))
t.lc + geom_line(fill=NA, size=1) + labs(x="Income Level") + labs(y="Travel Costs (minutes)") + ylim(0, 120) +
  ggtitle("Trip-level travel cost by trip purposes income groups") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))
```

### Tampa Bay
```{r, echo=FALSE}
setwd("~/tci")
load("output/NHTS09/TampaBay/tcost.RData")
```

**Tamba Baydata description**
```{r}
# Number of household
nrow(tcost.hh)

# Number of household by household size
tcost.hh.hhsiz.freq <- tcost.hh %>%
                 group_by(HHSIZ) %>%
                 summarise(freq=n())
tcost.hh.hhsiz.freq
# Number of trips by income groups and and trip purposes 
tcost.hh.tpurp.inc.freq <- tcost.trip %>%
                           group_by(TripPurpose, inc.level) %>%
                           summarise(freq=n()) %>%
                           as.data.frame()
tcost.hh.tpurp.inc.freq
```


**Tampa Bay: density_tcost.hh_by_inc.png**

```{r}
# Plot trip cost by income groups 
pden.inc <- ggplot(tcost.hh, aes(x = tcost, colour=inc.level, group=inc.level)) +
            geom_density(fill=NA, size=1) + labs(x="Travel Costs (minutes)") + xlim(0, 360) +
            scale_colour_discrete(name = 'Income Level') + 
  ggtitle("Household-level travel cost by household size") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))

pden.inc

```


**Tampa Bay: density_tcost.hh_by_hhsiz.png**
```{r}
tcost.hh <- tcost.hh %>% 
  mutate(hhsiz.cat=cut(HHSIZ,
                       breaks=c(1, 2, 3, 4, 9),
                       labels=c("1", "2", "3", "4+"),   #allow alternative household grouping
                       include.lowest=T, right=F
  ))
pden.hhsiz <- ggplot(tcost.hh, aes(x = tcost, colour=hhsiz.cat, group=hhsiz.cat)) +
  geom_density(fill=NA, size=1) + labs(x="Travel Costs (minutes)") + xlim(0, 360) +
  scale_colour_discrete(name = 'Household Size') + 
  ggtitle("Household-level travel cost by household size") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))
pden.hhsiz
```

**Tampa Bay: boxplot_tcost.hh_by_tpurp.inc.png**

```{r}
tcost.hh.tpurp <- tcost.hh.tpurp %>% 
  mutate(inc.level = factor(inc.level, levels=Ic, labels=c("Low Inc", "Mid Inc", "High Inc")),
         TripPurpose = factor(TripPurpose, levels=Pr, labels=c("HBW", "HB Shopping", "HB Recreation", "HB Other"))
  )

boxp.tpurp_inc <- ggplot(tcost.hh.tpurp, aes(x=TripPurpose, y=tcost, fill=inc.level)) +
  geom_boxplot() + labs(y="Generalized Travel Costs (minutes)") + xlab("Trip Purpose") + ylim(0, 250)  +
  scale_fill_discrete(name = 'Income Level') + 
  ggtitle("Household-level travel cost by trip purposes income groups") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))
boxp.tpurp_inc
```

**Tampa Bay: lineplot_tcost.hh_by_tpurp.inc.png**
```{r}
t.lc <- ggplot(tcost.tpurp.inc, aes(x = inc.level, y = tcost.wtavg, colour=TripPurpose, group=TripPurpose))
t.lc + geom_line(fill=NA, size=1) + labs(x="Income Level") + labs(y="Travel Costs (minutes)") + ylim(0, 120) +
  ggtitle("Trip-level travel cost by trip purposes income groups") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))
```

### Salt Lake City
```{r, echo=FALSE}
setwd("~/tci")
load("output/NHTS09/SaltLakeCity/tcost.RData")
```

**Salt Lake City data description**
```{r}
# Number of household
nrow(tcost.hh)

# Number of household by household size
tcost.hh.hhsiz.freq <- tcost.hh %>%
                 group_by(HHSIZ) %>%
                 summarise(freq=n())
tcost.hh.hhsiz.freq
# Number of trips by income groups and and trip purposes 
tcost.hh.tpurp.inc.freq <- tcost.trip %>%
                           group_by(TripPurpose, inc.level) %>%
                           summarise(freq=n()) %>%
                           as.data.frame()
tcost.hh.tpurp.inc.freq
```


**Salt Lake City: density_tcost.hh_by_inc.png**

```{r}
# Plot trip cost by income groups 
pden.inc <- ggplot(tcost.hh, aes(x = tcost, colour=inc.level, group=inc.level)) +
            geom_density(fill=NA, size=1) + labs(x="Travel Costs (minutes)") + xlim(0, 360) +
            scale_colour_discrete(name = 'Income Level') + 
  ggtitle("Household-level travel cost by household size") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))

pden.inc

```


**Salt Lake City: density_tcost.hh_by_hhsiz.png**
```{r}
tcost.hh <- tcost.hh %>% 
  mutate(hhsiz.cat=cut(HHSIZ,
                       breaks=c(1, 2, 3, 4, 9),
                       labels=c("1", "2", "3", "4+"),   #allow alternative household grouping
                       include.lowest=T, right=F
  ))
pden.hhsiz <- ggplot(tcost.hh, aes(x = tcost, colour=hhsiz.cat, group=hhsiz.cat)) +
  geom_density(fill=NA, size=1) + labs(x="Travel Costs (minutes)") + xlim(0, 360) +
  scale_colour_discrete(name = 'Household Size') + 
  ggtitle("Household-level travel cost by household size") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))
pden.hhsiz
```

**Salt Lake City: boxplot_tcost.hh_by_tpurp.inc.png**

```{r}
tcost.hh.tpurp <- tcost.hh.tpurp %>% 
  mutate(inc.level = factor(inc.level, levels=Ic, labels=c("Low Inc", "Mid Inc", "High Inc")),
         TripPurpose = factor(TripPurpose, levels=Pr, labels=c("HBW", "HB Shopping", "HB Recreation", "HB Other"))
  )

boxp.tpurp_inc <- ggplot(tcost.hh.tpurp, aes(x=TripPurpose, y=tcost, fill=inc.level)) +
  geom_boxplot() + labs(y="Generalized Travel Costs (minutes)") + xlab("Trip Purpose") + ylim(0, 250)  +
  scale_fill_discrete(name = 'Income Level') + 
  ggtitle("Household-level travel cost by trip purposes income groups") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))
boxp.tpurp_inc
```

**Salt Lake City: lineplot_tcost.hh_by_tpurp.inc.png**
```{r}
t.lc <- ggplot(tcost.tpurp.inc, aes(x = inc.level, y = tcost.wtavg, colour=TripPurpose, group=TripPurpose))
t.lc + geom_line(fill=NA, size=1) + labs(x="Income Level") + labs(y="Travel Costs (minutes)") + ylim(0, 120) +
  ggtitle("Trip-level travel cost by trip purposes income groups") +
  theme(plot.title = element_text(face="bold", size=12, vjust=1))
```
